---
title: "Vanderbilt Surgery Analysis"
output: slidy_presentation
---

# Import Libraries

```{r}
library(tidyverse)
library(readxl)
```

# Import Data

```{r}
#Import Data
df  <- read_xlsx('Vanderbilt Univ Case Dataset.xlsx')
```

```{r}
#print first few rows
df %>%
  head()
```

## Inspect Dataframe

```{r}
#print df shape
df %>% 
  dim()

#print df rows, cols
df %>% 
  nrow()

df %>% 
  ncol()

df %>% 
  length()
```

## Pre-Processing

```{r}
# Convert categorical variable to factor
df <- df %>% 
  mutate(DOW = factor(DOW, levels = c("Mon", "Tue", "Wed", "Thu", "Fri")))
```

# Exploratory Data Analysis

```{r}
#summarize
df %>% 
  summary()
```

## Correlation Matrix

```{r}
# Correlation
# select only numeric features
cor_df <- df %>% 
        select(-c(SurgDate, DOW)) %>% 
        cor() %>% 
        as.data.frame() %>% 
        round(2)

cor_df
```

## Correlation

```{r}
#inspect Correlation values
cor_df['Actual', 'T - 1'] #.96
cor_df['Actual', 'T - 3'] #.91
```

```{r}
#tabulate correlation
cor_df %>% 
  group_by(row.names(cor_df)) %>% 
  summarise(Cor = Actual) %>% 
  arrange(-Cor)
```

# ggplot Visualizations

```{r}
library(ggplot2)
# avg, std of actual surgeries by DOW
df %>% 
  group_by(DOW) %>% 
  summarise(N = n(),
            AVG = mean(Actual),
            SD = sd(Actual))
```

## Histogram

```{r}
df %>% 
  ggplot(aes(x = Actual)) + 
  geom_histogram(bins = 30, fill = "gray", color = "black") +
  theme_minimal() +
  labs(title = "Frequency of Daily Surgery Counts",
       x = "",
       y = "")
```

## Remove Outliers

```{r}
df %>% 
  filter(Actual > 50) %>% 
  ggplot(aes(x = Actual)) + 
  geom_histogram(bins = 30, fill = "gray", color = "black") +
  theme_minimal() +
  labs(title = "Frequency of Daily Surgery Counts",
       x = "",
       y = "")
```

## Bar Plot

```{r}
# Plot surgeries by DOW
df %>% 
  group_by(DOW) %>% 
  summarise(AVG = mean(Actual)) %>% 
  ggplot(aes(x = DOW, y = AVG)) + 
  geom_bar(stat = "identity", fill = "gray", color = "black") + 
  theme_minimal() + 
  labs(title = "Average Number of Surgeries by Day of Week",
       x = "",
       y = "") + 
  geom_text(aes(label = round(AVG, 0)), vjust = -0.5, hjust = 0.5, size = 4, position = 'stack')
```

## Box Plot

```{r}
#box plot
df %>% 
  ggplot(aes(x = DOW, y = Actual)) + 
  geom_boxplot(fill = "gray") + 
  theme_minimal() + 
  labs(title = "Average Number of Surgeries by Day of Week",
       x = "",
       y = "")
```

## Remove Outliers

```{r}
#box plot
df %>% 
  filter(Actual > 50) %>% 
  ggplot(aes(x = DOW, y = Actual)) + 
  geom_boxplot(fill = "gray") + 
  theme_minimal() + 
  labs(title = "Average Number of Surgeries by Day of Week",
       x = "",
       y = "")
```

# Statistics

```{r}
# stats table
df %>% 
  pivot_longer(cols = starts_with("T"), names_to = "T", values_to = "scheduled") %>% 
  mutate(T = substr(T, nchar(T) - 1 , nchar(T))) %>% 
  group_by(T) %>% 
  summarise(Max = max(scheduled),
            Mean = mean(scheduled),
            SD = sd(scheduled)
            )
```

# ANOVA Testing

- Null Hypothesis: Total surgical case volume does not differ based on the day of week 
- Alternative Hypothesis: Total surgical case volume differs based on the day of week

```{r}
library(caret)
model <- lm(Actual ~ DOW, data = df)
anova(model)
```

## ANOVA Results

- The p value obtained from ANOVA analysis is significant (p \< 0.05)

- Therefore, we can conclude that there is a significant difference among treatments.

- We reject the null hypothesis and accept the alternative hypothesis that total surgical case volume differ by DOW

## Model Summary

```{r}
summary(model)
```

# Forecasting (Linear Regression)

- Dependent variable: Actual surgeries
- Independent variables: T-28 to T-1

$$
y = a_1(T-28) + a_2(T-21)+ ... + a_x(T - 1) + b
$$

## Baseline Model 

```{r}
# remove outliers
df <- df %>% 
  filter(Actual > 50)
```

```{r}
# define models

#restrict df to variables we're using
df <- df %>% 
  select(starts_with(c("T", "A")))

#baseline: t-28 to t-1
m1 <- lm(Actual ~ ., data = df)
```

## 3 Days Prior

```{r}
#3 days prior: t-28 to t-4
m2 <- lm(Actual ~ `T - 28`+
                  `T - 21`+
                  `T - 14`+
                  `T - 13`+
                  `T - 12`+
                  `T - 11`+
                  `T - 10`+
                  `T - 9` +
                  `T - 8` +
                  `T - 7` +
                  `T - 6`+
                  `T - 5`+
                  `T - 4`,
         data = df)
summary(m2)
```

## 7 Days Prior

```{r}
#7 days prior: t-28 to t-8
m3 <- lm(Actual ~ `T - 28`+
                  `T - 21`+
                  `T - 14`+
                  `T - 13`+
                  `T - 12`+
                  `T - 11`+
                  `T - 10`+
                  `T - 9` +
                  `T - 8`,
         data = df)
summary(m3)
```

## Predictions

```{r}
#make predictions
p1 <- m1 %>% 
  predict(df)

p2 <- m2 %>% 
  predict(df)

p3 <- m3 %>% 
  predict(df)
```

## How close are the predictions to the actual values?

```{r}
#check average daily error
data.frame(Actual = df$Actual, #actual number of surgeries
           `Model 1 Predictions` = p1, #Model predictions
           Difference = abs(df$Actual - p1)) %>% #absolute value of (actual - predicted)
  summarise(`Model 1: Average Daily Error` = round(mean(Difference))) # aggregate average and round
```

```{r}
data.frame(Actual = df$Actual,
           `Model 2 Predictions` = p2,
           Difference = abs(df$Actual - p2)) %>% 
  summarise(`Model 2: Average Daily Error` = round(mean(Difference)))
```

```{r}
data.frame(Actual = df$Actual,
           `Model 3 Predictions` = p3,
           Difference = abs(df$Actual - p1)) %>% 
  summarise(`Model 3: Average Daily Error` = round(mean(Difference)))
```
